Efficient Calculation of Polynomial Features on Sparse Matrices

### 2016-11-01 paper/Final/sparse_poly.tex
fixed bibliography

\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{bm}
\usepackage{clrscode3e}
\usepackage{esvect}
\usepackage[T1]{fontenc}
\usepackage{hyperref}
\usepackage{mathtools}
\usepackage{times}
\usepackage{url}

@@

\author{
  Nystrom, Andrew\\
  \texttt{awnystrom@gmail.com}
  \and
  Hughes, John\\
  \texttt{jfh@cs.brown.edu}
}

@@

%\newcommand{\fix}{\marginpar{FIX}}
%\newcommand{\new}{\marginpar{NEW}}
\newcommand{\matr}[1]{\bm{#1}}

@@

\begin{abstract}
We provide an algorithm for polynomial feature expansion that operates directly on and produces a compressed sparse row matrix without any densification.
For a vector of dimension $D$ and density $d$, the algorithm has time and space complexity $O(d^kD^k)$ where $k$ is the polynomial order, which is an improvement by a factor $d^k$ over the standard method.
\end{abstract}

% this is a test-change for Spike to see whether Git seems to work for him. Ignore.

@@

\section{Preliminaries}
A matrix is uppercase and bold, like this: $\bm{A}$.
In this the context of this paper, row vectors are discussed, but not collum vectors.
Therefore, the distinction between the two is unimportant.
A row vector $\bm{a}_i$ is the $i^{th}$ row of $\bm{A}$ and $\bm{a}$, without a subscript, is a vector not necessarily related to $\bm{A}$.

A compressed sparse row (CSR) matrix representation of $\bm{A}$ consists of three vectors: $\bm{c}$, $\bm{d}$, and $\bm{p}$.
$\bm{c}$ and $\bm{d}$ hold the column indices and data values, respectively, of all nonzero elements of $\bm{A}$.
The values of $\bm{p}$ index $\bm{c}$ and $\bm{d}$ and is itself indexed by rows of $\bm{A}$ such that the nonzero columns of $\bm{a}_i$ are $\bm{c}_{\bm{p}_i:\bm{p}_{i+1}}$ and the nonzero data elements are $\bm{d}_{\bm{p}_i:\bm{p}_{i+1}}$.

@@

Note that in the algorithm adapted for CSR matrices, the counter is not reset for each row.
This is because there is only one $\bm{c}^k$ and $\bm{d}^k$ vector for all of $\bm{A}^k$, not one for each row.

However, note line \ref{li:set_ck}.
Given the arbitrary columns involved in a polynomial feature of $\bm{A}_i$, there is no way to determine what the corresponding column of $\bm{A}^k_i$.
In order to achieve a general algorithm, we require a mapping from columns of $\bm{A}$ a column of $\bm{A}^k$.
If there are $D$ columns of $\bm{A}$ and $\binom{D}{k}$ columns of $\bm{A}^k$, this can be accomplished by a bijective mapping of the following form:

@@

\section{Construction of Mapping}
For the second degree case, we seek a map from matrix indices $(i, j)$ (with $0 \le i < j < D$ ) to numbers $f(i, j)$ with $0 \le f(i, j) < \frac{D(D-1)}{2}$, one that follows the pattern indicated by 
\begin{align}
\begin{bmatrix}
x & 0 & 1 & 3 \\
x & x & 2 & 4 \\
x & x & x & 5 \\
x & x & x & x
\end{bmatrix}
\label{eq:4x4mat}
\end{align}
where the entry in row $i$, column $j$, displays the value $f(i, j)$. We let $T_2(n) = \frac{1}{2} n(n+1)$ 
be the $n$th triangular number; then in Equation~\ref{eq:4x4mat}, column $j$ (for $j > 0$) contains entries with  
$T_2(j-1) \le e < T_2(j)$; the entry in the $i$th row is just $i + T_2(j-1)$. Thus we have
$
f(i, j) 
= i + T_2(j-1) =  \frac{1}{2}(2i + j^2-j).$
For instance, in column $j = 2$ in our example (the \emph{third} column), the entry in row $i = 1$ is 
$i + T_2(j-1) = 1 + 1 = 2$.

@@

\subsubsection{Polynomial Features}
For polynommial features, we seek a map from matrix indices $(i, j)$ (with $i \le j$ and $0 \le i < D$) to numbers $g(i, j)$ with $0 \le f(i, j) < \frac{D(D+1)}{2}$, one that follows the pattern indicated by 
\begin{align}
\begin{bmatrix}
 0 & 1 & 3 & 6 \\
 x & 2 & 4 & 7\\
 x & x & 5 & 8 \\
 x & x & x & 9
\end{bmatrix}
\label{eq:4x4mat-poly}
\end{align}
i.e., essentially the same task as before, except that the diagonal is included. One can regard all but the last column of entries in Equation~\ref{eq:4x4mat-poly} as corresponding to the entries in Equation~\ref{eq:4x4mat}, but shifted to the left. Thus the formula for $g(i, j)$ is simply the formula for $f$, shifted by 1, i.e., 
\begin{align}
g(i, j) &= f(i, j+1)  \\
&=  \frac{2i + (j+1)^2-(j+1)}{2}\\
&=  \frac{2i + j^2 + j + 1)}{2}.
\end{align}
Alternatively, we can write this as
\begin{align}
g(i, j) &= i + T_2(j),
\end{align}
\noindent
and get the same result.

\subsubsection{Higher dimensions}
To handle three-way interactions, we need to map triples of indices in a 3-index array to a flat list, and similarly for higher-order interactions.

For three indices, $i,j,k$, with $i < j < k$ and $0 \le i,j,k < D$, we have a similar recurrence. Calling the mapping $h$, we have 
\begin{align}
h(i,j,k) = i + T_2(j-1) + T_3(k-2);
\end{align}
if we define $T_1(i) = i$, then this has the very regular form
\begin{align}
h(i,j,k) =  T_1(i) + T_2(j-1) + T_3(k-2);
\end{align}
and from this, the generalization to higher dimensions is straightforward. The formulas for ``higher triangular numbers'', i.e., those defined by
\begin{align}
T_k(n) &= \sum_{i=1}^n T_{k-1}(n)
\end{align}
for $k > 1$ can be determined inductively. For $k = 3$, the result is 
\begin{align}
T_3(n) &= \sum_{i=1}^n T_{2}(n)\\
&= \frac{n^3 + 3n^2 + 2n}{6},
\end{align}
so that the formula for 3-way interactions, with zero-based indexing, becomes 
\begin{align}
h(i, j, k) &= 1 + (i-1) + \frac{(j-1)j}{2} + \\
& \frac{(k-2)^3 + 3(k-2)^2 + 2(k-2)}{6}. 
\end{align}
\subsubsection{Higher-dimension polynomial features}
For the case where we include the diagonal in higher dimensions, we must shift $j$ by $1$, $k$ by $2$, and so on, and the formula becomes
\begin{align}
\ell(i,j,k) =  T_1(i) + T_2(j) + T_3(k),
\end{align}
with analogous formulas for higher degree polynomial interactions.

\subsection{Inversion}
We not only want to be able to compute the ``flat index'' $f(i,j)$ corresponding to a particular index pair $(i, j)$, we'd like, given a value $N = f(i, j)$, to be able to determine $i$ and $j$. We know, given $N$, that $f(0, j) \le N \le f(j,j-1)$, i.e., that 
\begin{align}
\frac{j^2 - 3j + 2}{2} &\le N \le \frac{2(j-1) + j^2 - 3j + 2}{2}\\
{j^2 - 3j + 2} &\le 2N \le {j^2 - j}.
\end{align}
If we find the (real number) value $j_{*}$ for which $j_{*}^2 - j_{*} = 2N$, then we know that $j = \lfloor j_{*}\rfloor$. Finding $j_{*}$ amounts to solving a quadratic, whence
\begin{align}
j_{*} = \frac{1 \pm \sqrt{1 + 8N}}{2}.
\end{align}
Clearly the larger value, 
\begin{align}
j_{*} = \frac{1 + \sqrt{1 + 8N}}{2},
\end{align}
is the one we seek; we set $j = \lfloor j_{*}\rfloor$, and compute $i = N - \frac{j^2 - j}{2}$. 
In practice, repeatedly computing square roots may be annoying; instead, when we compute $N = f(i,j)$, we can record the values of $i$ and $j$ along with the interaction features, making inversion a simple matter of looking up these values.

Similarly, for polynomial features, to invert 
\begin{align}
g(i, j) &=  \frac{2i + j^2 + j + 1)}{2}
\end{align}
we know that if $N = g(i,j)$, then $g(0,j) \le N \le g(j,j)$, which tells us that 
\begin{align}
2N \le j^2 + 3j + 1, \text{ so}\\
0 \le j^2 + 3j + (1-N^2)
\end{align}. 
Solving for $j$ when this last inequality is an equality, we get 
\begin{align}
j_{*} = \frac{-3 \pm \sqrt{ 5 - N^2 }}{2}; 
\end{align}
Once again, we set $j = \lfloor j_{*}\rfloor$, and compute $i = N - \frac{j^2 + j + 1}{2}$.

Inverting the higher-order formulas, which involve cubics, quartics, etc., is infeasible, but storing the inverse map as described above may be practical, especially for sparse feature sets.

\section{Final CSR Expansion Algorithm}
With the mapping from columns of $\bm{A}$ to a column of $\bm{A}^k$, we can now write the final form of the innermost loop of the algorithm from \ref{sec:final-algo}.
Let the mapping for $k=2$ be denoted $PolyCol_2$.
Then the innermost loop becomes:

\begin{codebox}
\footnotesize
    \zi         \For $c_2 \gets c_1 \To |\bm{c}_i|-1$ \Do
    \zi             $j_0 \gets \bm{c}_{c_0}$
    \zi             $j_1 \gets \bm{c}_{c_1}$
    \zi             $c_p \gets PolyCol_2(j_0, j_1)$
    \zi             $\bm{d}^k_{n} \gets \bm{d}_{c_0} \cdot \bm{d}_{c_1}$
    \zi             $\bm{c}^k_{n} = c_p$
    \zi             $n \gets n + 1$
                \End
\end{codebox}

\section{Time Complexity}
\subsection{Analytical}
Calculating $k$-degree polynomial features via our method for a row $\bm{a}_i$ of a matrix of dimensionality $D$ and density $d$ requires $\binom{dD}{k}$ (with repetition) products, assuming the nonzero elements of $\bm{A}$ are uniformly distributed.
The complexity of the algorithm, for fixed $k \ll dD$, is therefore
\begin{align}
O\left(\binom{dD+k-1}{k}\right) & = O\left(\frac{(dD+k-1)!}{k!(dD-1)!}\right)\\
& = O\left(\frac{(dD+k-1)(dD+k-2) \dots (dD)}{k!}\right)\\
& = O\left((dD+k-1)(dD+k-2) \dots (dD)\right) \mbox{ for } k \ll dD\\
& = O\left(d^kD^k\right)
\end{align}

\subsection{Empirical}
\section{Conclusion}

\bibliography{sparse_poly}
\bibliographystyle{iclr2016_workshop}

>>>>

\usepackage{amssymb}
\usepackage{bm}
\usepackage{clrscode3e}
\usepackage{float}
\usepackage{mathtools}

%icrl 2017
\usepackage{iclr2017_conference,times}
\usepackage{hyperref}
\usepackage{url}

@@

\author{
  Nystrom, Andrew \\
  \texttt{nystrom@google.com}
  \and
  Hughes, John \\
  \texttt{jfh@cs.brown.edu}
}

@@

\newcommand{\fix}{\marginpar{FIX}}
\newcommand{\new}{\marginpar{NEW}}

@@

\begin{abstract}
We provide an algorithm for polynomial feature expansion that operates directly on and produces a compressed sparse row matrix without any densification.
For a vector of dimension $D$, density $d$, and degree $k$ the algorithm has time and space complexity $O(d^kD^k)$ where $k$ is the polynomial order, which is an improvement by a factor $d^k$ over the standard method.
\end{abstract}

@@

\section{Preliminaries}
A matrix $\bm{A}$ is uppercase and bold.
A row vector $\bm{a}_i$ has a subscript and is the $i^{th}$ row of $\bm{A}$, whereas $\bm{a}$, without a subscript, is a vector not necessarily related to $\bm{A}$.

A compressed sparse row (CSR) matrix representation of $\bm{A}$ consists of three vectors: $\bm{c}$, $\bm{d}$, and $\bm{p}$.
$\bm{c}$ and $\bm{d}$ hold the column indices and data values, respectively, of all nonzero elements of $\bm{A}$.
The values of $\bm{p}$ index $\bm{c}$ and $\bm{d}$ and is itself indexed by rows of $\bm{A}$ such that the nonzero columns of $\bm{a}_i$ are $\bm{c}_{\bm{p}_i:\bm{p}_{i+1}}$ and the nonzero data elements are $\bm{d}_{\bm{p}_i:\bm{p}_{i+1}}$.
Since only nonzero elements of each row are held, the dimensionality must also be stored since it cannot be implicitly gotten from the length of $\bm{c}$.

Scalers, vectors, and matrices are often referenced with the superscript $k$.
This is not to be interpreted as an exponent, but to indicate that it is the analogous aspect of that which procedes it, but in its polynomial expansion form.
For example, $\bm{c}^k$ is the vector that holds columns for nonzero values in $\bm{A}$'s polynomial feature expansion CSR representation.

For simplicity, all polynomial expansions are assumed to be of the second degree.
We do provide an algorithm for third degree expansions and derive the big O time complexity of the general case.

@@

The crux of the problem can be found on line \ref{li:set_ck}.
Given the arbitrary columns involved in a polynomial feature of $\bm{A}_i$, there is no way to determine what the corresponding column of $\bm{A}^k_i$.
We cannot simply reset a counter for each row as we did in the dense algorithm.
This is because only columns corresponding with nonzero values are stored.
Any time a column is implicitly skipped that would have held a zero value, the counter would err.

In order to achieve a general algorithm, we require a mapping from columns of $\bm{A}$ a column of $\bm{A}^k$.
If there are $D$ columns of $\bm{A}$ and $\binom{D}{k}$ columns of $\bm{A}^k$, this can be accomplished by a bijective mapping of the following form:

@@

\section{Construction of Mapping}
Within this section, $i$, $j$, and $k$ denote column indices.
For the second degree case, we seek a map from matrix indices $(i, j)$ (with $0 \le i < j < D$ ) to numbers $f(i, j)$ with $0 \le f(i, j) < \frac{D(D-1)}{2}$, one that follows the pattern indicated by 
\begin{align}
\begin{bmatrix}
x & 0 & 1 & 3 \\
x & x & 2 & 4 \\
x & x & x & 5 \\
x & x & x & x
\end{bmatrix}
\label{eq:4x4mat}
\end{align}
where the entry in row $i$, column $j$, displays the value $f(i, j)$. We let $T_2(n) = \frac{1}{2} n(n+1)$ 
be the $n$th triangular number; then in Equation~\ref{eq:4x4mat}, column $j$ (for $j > 0$) contains entries with  
$T_2(j-1) \le e < T_2(j)$; the entry in the $i$th row is just $i + T_2(j-1)$. Thus we have
$
f(i, j) 
= i + T_2(j-1) =  \frac{1}{2}(2i + j^2-j).$
For instance, in column $j = 2$ in our example (the \emph{third} column), the entry in row $i = 1$ is 
$i + T_2(j-1) = 1 + 1 = 2$.

@@

\section{Final CSR Expansion Algorithm}
With the mapping from columns of $\bm{A}$ to a column of $\bm{A}^k$, we can now write the final form of the innermost loop of the algorithm from \ref{sec:final-algo}.
Let the mapping for $k=2$ be denoted $PolyCol^2$.
Then the innermost loop becomes:

\begin{codebox}
\footnotesize
    \zi         \For $c_2 \gets c_1 \To |\bm{c}_i|-1$ \Do
    \zi             $j_0 \gets \bm{c}_{c_0}$
    \zi             $j_1 \gets \bm{c}_{c_1}$
    \zi             $c_p \gets PolyCol^2(j_0, j_1)$
    \zi             $\bm{d}^k_{n} \gets \bm{d}_{c_0} \cdot \bm{d}_{c_1}$
    \zi             $\bm{c}^k_{n} = c_p$
    \zi             $n \gets n + 1$
                \End
\end{codebox}

The algorithm can be generalized to higher degrees by simply adding more nested loops and using higher order mappings.

\section{Time Complexity}
\subsection{Analytical}

Calculating $k$-degree polynomial features via our method for a vector of dimensionality $D$ and density $d$ requires $\binom{dD}{k}$ (with repetition) products.
The complexity of the algorithm, for fixed $k \ll dD$, is therefore
\begin{align}
O\left(\binom{dD+k-1}{k}\right) & = O\left(\frac{(dD+k-1)!}{k!(dD-1)!}\right)\\
& = O\left(\frac{(dD+k-1)(dD+k-2) \dots (dD)}{k!}\right)\\
& = O\left((dD+k-1)(dD+k-2) \dots (dD)\right) \mbox{ for } k \ll dD\\
& = O\left(d^kD^k\right)
\end{align}

\subsection{Empirical}
To demonstrate how our algorithm scales with the density of a matrix, we compare it to the traditional polynomial expansion algorithm in the popular machine library scikit-learn \cite{scikit-learn} in the task of generating second degree polynomial expansions.
Matrices of size $100 \times 5000$ were randomly generated with densities of $0.2$, $0.4$, $0.6$, $0.8$, and $1.0$.
Thirty matrices of each density were randomly generated, and the mean times (gray) of each algorithm were plotted.
The red or blue width around the mean marks the third standard deviation from the mean.

The standard algorithm's runtime stays constant no matter the density of the matrix.
This is because it does not avoid products that result in zero, but simply multiplies all second order combinations of features.
Our algorithm scales quadratically with respect to the density.
If the task were third degree expansions rather than second, the plot would show cubic scaling.

The fact that our algorithm is approximately $6.5$ times faster than the scikit-learn algorithm on $100 \times 5000$ matrices that are entirely dense is likely an language implementation difference.
What matters is that the time of our algorithm increases quadratically with respect to the density in accordance with the big O analysis.

\begin{figure}[H]
    \centering
    \includegraphics[scale=0.6]{density_vs_time}
    \caption{This plot shows how our algorithm (bottom) scales with the density of a matrix compared to a traditional polynomial feature expansion method (top).
             The task was a second degree expansion, which is why the time of our algorithm scales quadratically with the density.}
    \label{fig:benchmark}
\end{figure}

\section{Conclusion}
We have developed an algorithm for performing polynomial feature expansions on CSR matrices that scales polynomially with respect to the density of the matrix.
While the areas within machine learning this work touches is not the flavor of the day, they are workhorses of industry and will likely save much coal from being unnecessarily burnt.

\bibliography{sparse_poly}
\bibliographystyle{iclr2017_conference}
